Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-10-11 analysis/paper/paper.Rmd
added sentences to method section

# pottery shape analysis 
- review about pottery shape analysis recently

@@

Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan. Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period with the onset of the European presence in the 17th century. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. The Upper Culture Layer experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan. In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a large settlement site.

@@

## geometric morphometric approach

The drawings of pottery with half cross-section from rim to body were imported into Inkscape software for outline digitalization. All half cross-section images were copied and flipped to create enclosed images in order to be ran in the R environment. The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts. The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction. The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances. Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation. Outline analyses were carried out using the package Momocs in the software R.

>>>>

# Geometric Morphometric approach: shape analysis for pottery specialization

Geometric morphometry methods (GMM) has been widely applied recently in archaeology to study the morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from taking linear measurements of objects as variables, geometric morphometrics methods use geometric coordinates of morphological structures as variables to specify the shapes [@Adams2004; @Lawing2010]. According to different focuses, there are two common methods used to capture and analyze the shapes of objects, landmark analysis and outline analysis [@Adams2004]. Landmark analysis

- related to the issue of specialization

@@

Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan. Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period  by the European presence in the 17th century. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. The Upper Culture Layer experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan. In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a large settlement site.

@@

## Outline analysis approach

The scanned pottery drawings for reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original file. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015). Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013]. Here we used 13 harmonics to describe pottery shapes. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction. The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances. Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups. The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.
